Smart Spreading Factor Assignment for LoRaWANs

### 2019-01-17 paper.tex
grammer

\par LoRa can adjust data rate by spreading symbols within a fixed channel bandwidth. This enables tradeoff between receive sensitivity and transmission time on air \cite{7803607}. Simultaneous same SF transmissions are prone to collision, however, different SF transmissions in the same channel are orthogonal to each other, thus SF assignment is crucial for overall network performance. In this work, a LoRa discrete event simulator is developed to study different LoRa SF assignment strategies. We also proposed an SF assignment scheme to increase network performance with the help of machine learning methods and we verified our proposal using the simulator.

@@

\subsection{Spreading Factor Assignment Issue}
\par Simultaneous different SF transmissions are orthogonal to each other up to some extent. Which means, a LoRa GW can simultaneously receive multiple transmissions with different SFs. However, simultaneous transmissions with same SF may not be received by GW due to collision. For this reason, SF assignment of nodes is crucial for network performance.

\par In a LoRaWAN network, initially, a node is not aware of how far it is from a GW. However, a node can guess the distance from a GW by observing received signal power of a downlink transmission. If a downlink transmission received signal power is too high, then node can decrease its next transmission SF to decrease power consumption. This SF assignment method is called lowest possible SF assignment scheme for the rest of the paper. Lowest possible SF assignment scheme is commonly used in LoRaWAN deployments. GW can request from a node to decrease its SF or transmit power.

\par In Figure \ref{fig:collision}, a LoRaWAN network deployed with a single GW is illustrated. Different color rings represent achievable range of different SFs from the GW and different color circles represent selected SF of the nodes. Most of the end devices which close to the GW will fall into lowest SF (SF7) area section. End devices which close to the GW will probably select lowest SF all the time. This results a lot of collisions between same SF transmissions. Number of collisions increases while number of end devices close to the GW increases.

@@

\section{LoRaWAN} \label{LoRaWAN}
\par LoRa has an open standard medium access control (MAC) layer protocol called LoRaWAN which designed for large scale LoRa networks considering well known LPWAN challenges and their best practice solutions. LoRaWAN is developed and maintained by LoRa Alliance. LoRa Alliance is an open, non-profit organization dedicated to standardization of LoRaWAN. LoRaWAN is based on pure ALOHA medium access which means end nodes do not check whether the channel is free or not before transmission, accepting the possibility of a collision. A typical LoRaWAN network consist of following three network entities.

\subsection{End node}
\par LoRaWAN end node (EN) is a low power embedded device that only communicates to GWs. LoRaWAN standard defines three classes for end devices which are Class A, Class B, and Class C. Different classes provides LPWAN solutions to different applications and deployments. Class A end nodes generate uplink transmission at any time and only receives a period of time after uplink transmission. Class B end nodes extend Class A behavior by adding scheduled receive windows for downlink transmission. Receive window is synchronized using a beacon packet transmitted by GWs. Class C end nodes extend Class A behavior by keeping receive window open all the time except uplink transmission. This provides Class C end nodes to low latency downlink communication, but it requires more power consumption. In this paper, we focused on Class A end devices since Class A behavior provides the lowest power consumption.

\subsection{Gateway}
\par LoRaWAN gateway (GW) is a device that communicates to end nodes. A typical GW can receive from multiple channels at the same time. GWs are typically connected to power grid, so power consumption of a GW is insignificant in most of the deployments.

@@

\section{Other Related Works} \label{Other Related Works}
\par The literature related to the work presented in this paper has started growing recently. LPWAN technologies and especially LoRa attracted researchers’ attentions lately. We listed some of these works which studies LoRaWAN and LoRa SF.

@@

\par In \cite{8090518}, another LoRaWAN ns-3 simulator is presented. Authors introduced an error model for determining range as well as interference between multiple simultaneous LoRa transmissions. Their simulator supports LoRaWAN Class A end devices, multiple GWs, both upstream and downstream confirmed messages. Their results show that allocating network parameters to end devices is hugely important for the performance of LoRaWAN networks.

\par In \cite{8267219}, the authors studied imperfect orthogonality between different LoRa SF transmissions. The authors state that a LoRa transmission can be interfered even between different SF transmissions when power of the interfering signal significantly overcomes the reference signal. Their experimental results show that this power difference is around 16 dB and this power difference can be seen when an interferer is close to a receiver or sum of multiple interferer energy can create this power difference.

\par In \cite{8430542}, the authors investigated the impact of interference caused by simultaneous LoRa transmissions with same and different SFs. They derived aggregated co-SF and inter-SF interference power SIR distributions to capture the coverage distance from the GW for modeling interference in multiple GWs scenarios. Their results show that transmission among different SFs can cause a significant impact in high-density LoRaWAN networks.

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} can be solved by forcing some of the close nodes to select higher SF even they can able to communicate with lower SF. This may prevent collisions due to the orthogonality of different SF as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, how network should assign which SF to which node becomes an important question. Increasing a node's SF should be done carefully since higher SF means higher time on air and higher time on air is more prone to collision with other high SF transmissions. In multiple GW scenarios, this approach may increase the collisions with nodes in other GW's range. Thus, extra care should be taken for nodes in intersection area of GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes' SFs requires different rules. For this reason, we proposed a machine learning based SF assignment approach to decrease same SF transmission collisions. This technique works by learning the transmission behavior of the nodes in a network. A NS can keep track of successful uplink transmissions and their SFs. NS can also keep track of some of the collided transmission if header part of the packet is not interfered at location of the GW. However, NS cannot keep track of transmissions with receive power lower than the GW sensitivity. Using these obtained information, NS can train a classifier for predicting future transmission result for specific node and specific SF. Using this prediction model NS can assign the SF to nodes considering the collision probability.

@@

\par We have been able to generate mass amount of LoRaWAN transmission records for different topologies using the simulator. Data set size is directly proportional to simulation duration. Thus, increasing simulation duration, improves the prediction accuracy. In real world deployments, NS can keep track of all the transmission and it can create a classifier daily basis, then GWs can request from nodes to use their new SFs.

@@

\par In the simulator, only LoRaWAN Class A devices are covered. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet randomly according to Poisson interval for given packet rate input parameter. In the simulator, downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictates duty cycle limit for all devices including GWs.

@@

\par In the simulator, it is assumed that nodes always select maximum allowed transmit power, which is 14 dBm for European ISM band regulation \cite{lorawan.regional.parameters}. Different channel transmissions are independent from each other. But, in this work we focus on SF orthogonality. Thus, we only utilized single channel transmissions in the simulator.

@@

\par In our simulator, it is assumed that there is no other technology interference in the network except LoRa interference. To exploit imperfect orthogonality of different SF transmission, simulator should calculate the effect of different SF transmissions to each other. In simulator, following signal to interference plus noise ratio (SINR) threshold matrix from \cite{goursaud:hal-01231221} is used.

\par To decide if a referenced signal is interfered at receiver by a interferer signal, SINR threshold matrix is used. $T_{i,j}$ is the needed SINR margin in dB unit between referenced signal with SF = i and interferer signal with SF = j to correctly decode the referenced signal. If there are more than one interferer, referenced signal must satisfy the margin for cumulative sum of all interfering signal received powers for each SF \cite{7996384}. Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}.

@@

\par In Figure \ref{fig:sf_pdr}, PDR plot of different SF assignment is shown. Randomly generated network topology radius is set to 3000 meters and number of GWs is set to 1. Increasing SFs, increases time on air. This increases the number of collisions thus decreases the PDR. Higher SFs effected exponentially while number of nodes increases. Since network topology radius is quite short, all SFs can reach to the GW.

@@

\par In Figure \ref{fig:prediction_pdr}, result of proposed technique is presented. PDR plot of lowest SF scheme and machine learning prediction schemes are shown in same figure. Randomly generated network topology radius is set to 5000 meters and number of GWs is set to 3. Prediction model needs nodes location and three GWs are enough to roughly locate nodes positions by triangulation. In Table \ref{table:prediction_pdr}, PDR values for different topology radius are presented. Both prediction SVM and DTC SF schemes give better PDR than lowest SF schemes when number of nodes increases. Increasing number of nodes, increases number of interference. Prediction schemes  improve network performance when LoRa interference is high. Also, prediction schemes give better results when nodes are deployed close to the GW, since nodes have margin to increase their SFs when deployed close to the GW. If a node is far away from the GW, then prediction schemes cannot increase the SF to avoid interference since the SF is already high.

\par In Table \ref{table:prediction_accuracy}, prediction accuracy of SVM and DTC schemes are presented for different network topology radius and number of nodes. Prediction accuracy is not directly proportional to network PDR. Correct prediction of a interfered transmission may not increase the PDR but increases the accuracy metric. DTC gives better network PDR results than SVM, even overall SVM prediction accuracy is high. LoRaWAN simulation usually produces imbalanced dataset, DTC gives better accuracy for interfered transmission labels. We invite readers to experiment the simulation tool to review confusion matrices for different machine learning schemes.

@@

\section{Conclusion} \label{Conclusion}
\par In this paper, after a brief introduction about LPWAN technologies, we provide background information about LoRa and LoRaWAN. Then, we discuss about LoRa modulation basics and spreading factor assignment issue. We have implemented a discrete event simulator from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. We have shown how same SF transmission collisions can be avoided and we proposed machine learning based solution for collision with same SF transmissions. We illustrate simulation results for proposed machine learning SF assignment technique. Our simulation results show that, proposed technique can increase network performance for LoRaWAN networks especially nodes are deployed close to the GWs.

\par As for future works, transmit power optimization can be included to proposed machine learning scheme. In this paper, we assume that nodes always use maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmissions to more vulnerable to interference thus requires extra care. Also, other machine learning methods can be investigated for SF optimization. Reinforcement learning is a good candidate for this network optimization issue. Other transmission parameters such as node id and transmission time can be included to proposed scheme to improve performance.

>>>>

\par LoRa can adjust data rate by spreading symbols within a fixed channel bandwidth. This enables tradeoff between receive sensitivity and time on air of transmission \cite{7803607}. Simultaneous same SF transmissions are prone to collision, however, different SF transmissions in the same channel are orthogonal to each other, thus SF assignment is crucial for overall network performance. In this work, a LoRa discrete event simulator is developed to study different LoRa SF assignment strategies. Also an SF assignment scheme is proposed to increase network performance with the help of machine learning methods and the proposal is verified using the simulator.

@@

\subsection{Spreading Factor Assignment Issue}
\par Simultaneous different SF transmissions are orthogonal to each other up to some extent. Which means that, a LoRa GW can simultaneously receive multiple transmissions with different SFs. However, simultaneous transmissions with the same SF may not be received by the GW due to collision. For this reason, SF assignment of nodes is crucial for network performance.

\par In a LoRaWAN network, initially, a node is not aware of how far away it is from a GW. However, a node can guess the distance from a GW by observing received signal power of a downlink transmission. If received signal power of a downlink transmission is too high, then node can decrease its next transmission SF to decrease power consumption. This SF assignment method is called lowest possible SF assignment scheme for the rest of the paper. Lowest possible SF assignment scheme is commonly used in LoRaWAN deployments. GW can request from a node to decrease its SF or transmit power.

\par In Figure \ref{fig:collision}, a LoRaWAN network deployed with a single GW is illustrated. Different color rings represent achievable range of different SFs from the GW and different color circles represent selected SF of the nodes. Most of the end devices close to the GW will fall into lowest SF (SF7) area section. End devices close to the GW will probably select lowest SF all the time. This causes a lot of collisions between same SF transmissions. Number of collisions increases while number of end devices close to the GW increases.

@@

\section{LoRaWAN} \label{LoRaWAN}
\par LoRa has an open standard medium access control (MAC) layer protocol called LoRaWAN is which designed for large scale LoRa networks considering well known LPWAN challenges and their best practice solutions. LoRaWAN is developed and maintained by LoRa Alliance. LoRa Alliance is an open, non-profit organization dedicated to standardization of LoRaWAN. LoRaWAN is based on pure ALOHA medium access which means that end nodes do not check whether the channel is free or not before transmission, accepting the possibility of a collision. A typical LoRaWAN network consists of following three network entities.

\subsection{End node}
\par LoRaWAN end node (EN) is a low power embedded device that only communicates to GWs. LoRaWAN standard defines three classes for end devices which are Class A, Class B, and Class C. Different classes provide LPWAN solutions to different applications and deployments. Class A end nodes generate uplink transmission at any time and only receive a period of time after uplink transmission. Class B end nodes extend Class A behavior by adding scheduled receive windows for downlink transmission. Receive window is synchronized using a beacon packet transmitted by GWs. Class C end nodes extend Class A behavior by keeping receive window open all the time except uplink transmission. This provides Class C end nodes with low latency downlink communication, but it requires more power consumption. In this paper, only Class A end devices are considered since Class A behavior provides the lowest power consumption.

\subsection{Gateway}
\par LoRaWAN gateway (GW) is a device that communicates to end nodes. A typical GW can receive from multiple channels at the same time. GWs are usually connected to power grid, so power consumption of a GW is insignificant in most of the deployments.

@@

\section{Other Related Works} \label{Other Related Works}
\par The literature related to the work presented in this paper has started growing recently. LPWAN technologies and especially LoRa attracted researchers’ attentions lately. Some of these works which studies LoRa/LoRaWAN SF are listed.

@@

\par In \cite{8090518}, another LoRaWAN ns-3 simulator is presented. Authors introduced an error model for determining range as well as interference between multiple simultaneous LoRa transmissions. Their simulator supports LoRaWAN Class A end devices, multiple GWs, both upstream and downstream confirmed messages. Their results show that allocating network parameters to end devices is highly important for the performance of LoRaWAN networks.

\par In \cite{8267219}, the authors studied imperfect orthogonality between different LoRa SF transmissions. The authors state that a LoRa transmission can be interfered even between different SF transmissions when power of the interfering signal significantly overcomes the reference signal. Their experimental results show that this power difference is around 16 dB and this power difference can be seen when an interfering signal is close to a receiver or sum of interfering signals' energy can create this power difference.

\par In \cite{8430542}, the authors investigated the impact of interference caused by simultaneous LoRa transmissions with the same and different SFs. They derived aggregated co-SF and inter-SF interference power SIR distributions to capture the coverage distance from the GW for modeling interference in multiple GWs scenarios. Their results show that transmission among different SFs can cause a significant impact in high-density LoRaWAN networks.

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} is solved by forcing some of the close nodes to select higher SF even they are able to communicate with lower SF. This has potential to prevent collisions due to the orthogonality of different SF as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, distribution of SF among nodes becomes an important question. Increasing a node's SF should be done carefully since higher SF means longer time on air and longer time on air is more prone to collision with other high SF transmissions. In multiple GWs scenarios, this approach may increase the collisions with nodes in other GW's range. Thus, extra care should be taken for nodes in intersection area of the GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes' SFs requires different rules. For this reason, machine learning based SF assignment approach is proposed to decrease same SF transmission collisions. This technique works by learning the transmission behavior of the nodes in a network. A NS can keep track of successful uplink transmissions and their SFs. NS can also keep track of some of the collided transmission if header part of the packet is not interfered at location of the GW. However, NS cannot keep track of transmissions with receive power lower than sensitivity of the GW. Using these obtained information, NS can train a classifier for predicting future transmission result for specific node and specific SF. Using this prediction model NS can assign the SF to nodes considering the collision probability.

@@

\par It is possible to generate mass amount of LoRaWAN transmission records for different topologies using the simulator. Data set size is directly proportional to simulation duration. Thus, increasing simulation duration, improves the prediction accuracy. In real world deployments, NS can keep track of all the transmission and it can create a classifier daily basis, then GWs can request from nodes to use their new SFs.

@@

\par In the simulator, only LoRaWAN Class A devices are covered. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet randomly according to Poisson interval for given packet rate input parameter. In the simulator, downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictate duty cycle limit for all devices including GWs.

@@

\par In the simulator, it is assumed that nodes always select maximum allowed transmit power, which is 14 dBm for European ISM band regulation \cite{lorawan.regional.parameters}. Different channel transmissions are independent from each other. But, in this work we focus on SF orthogonality. Thus, only single channel transmissions are utilized in the simulator.

@@

\par In our simulator, it is assumed that there is no other technology interference in the network except LoRa interference. To exploit imperfect orthogonality of different SF transmission, simulator should calculate the effect of different SF transmissions to each other. In simulator, signal to interference plus noise ratio (SINR) threshold matrix from \cite{goursaud:hal-01231221} is used.

\par To decide if a referenced signal is interfered at receiver by an interfering signal, SINR threshold matrix is used. $T_{i,j}$ is the needed SINR margin in dB unit between referenced signal with SF = i and interfering signal with SF = j to correctly decode the referenced signal. If there are more than one interfering signal, referenced signal must satisfy the margin for cumulative sum of all interfering signal received power for each SF \cite{7996384}. Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}.

@@

\par In Figure \ref{fig:sf_pdr}, PDR plot of different SF assignment is shown. Randomly generated network topology radius is set to 3000 meters and number of GWs is set to 1. Increasing SFs, increases time on air. This increases the number of collisions thus decreases the PDR. High SF schemes gives poor PDR results while number of nodes increases. Since network topology radius is quite short, all SFs can reach to the GW.

@@

\par In Figure \ref{fig:prediction_pdr}, result of proposed technique is presented. PDR plot of lowest SF scheme and machine learning prediction schemes are shown in the same figure. Randomly generated network topology radius is set to 5000 meters and number of GWs is set to 3. Prediction model needs nodes location and three GWs are enough to locate nodes positions by triangulation. In Table \ref{table:prediction_pdr}, PDR values for various topologies with different radii are presented. Both prediction SVM and DTC SF schemes give better PDR than lowest SF schemes when number of nodes increases. Increasing number of nodes, increases number of interference. Prediction schemes  improve network performance when LoRa interference is high. Also, prediction schemes give better results when nodes are deployed close to the GW, since nodes have margin to increase their SFs when they are deployed close to the GW. If a node is far away from the GW, then prediction schemes cannot increase the SF to avoid interference since the SF is already high.

\par In Table \ref{table:prediction_accuracy}, prediction accuracy of SVM and DTC schemes are presented for different network topology radius and number of nodes. Prediction accuracy is not directly proportional to network PDR. Correct prediction of a interfered transmission may not increase the PDR but increases the accuracy metric. DTC gives better network PDR results than SVM, even overall SVM prediction accuracy is high. LoRaWAN simulation usually produces imbalanced dataset, DTC gives better accuracy for interfered transmission labels. We invite readers to experiment the simulation tool to review confusion matrices for SVM and DTC schemes.

@@

\section{Conclusion} \label{Conclusion}
\par In this work, after a brief introduction about LPWAN technologies, background information about LoRa and LoRaWAN is provided. Then, LoRa modulation basics and spreading factor assignment issue is  discussed. A discrete event simulator is implemented from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. Same SF transmission collisions are shown and machine learning based solution is proposed for collision with same SF transmissions. Simulation results for proposed machine learning SF assignment technique is illustrated. Simulation results show that, proposed technique can increase network performance for LoRaWAN networks especially nodes are deployed close to the GWs.

\par As for future works, transmit power optimization can be included in proposed machine learning scheme. In this paper, it is assumed that nodes always use maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmissions more vulnerable to interference thus requires extra care. Also, other machine learning methods can be investigated for SF optimization. Reinforcement learning is a good candidate for this network optimization issue. Other transmission parameters such as node id and transmission time can be included to proposed scheme to improve performance.
